---
title: "Analysis: COVID Effect on Employment"
author: "Group 2"
date: "`r Sys.Date()`"
knit: (function(input, ...) {rmarkdown::render(input, output_file="../paper/analysis.html")})
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Research Questions

* How has COVID affected the health of the retail industry, as measured by employment?
* How has retail fared relative to other industries?
* Retail needs to worry about who has money to spend - what has changed about who is      working and earning money?

## Libraries

``` {r libraries, echo=FALSE}
# Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
library(vtable)
library(fixest)
library(tidyverse)
library(lubridate)
```

## Read in Data

Bring in the processed data and store it in the schools variable. Filter on predominately bachelor degree schools. View the data.

```{r data}
question_1_data = read_csv("./data_processed/emp_count_by_retail_trade.csv")
question_2_data = read_csv("./data_processed/emp_count_by_industry.csv")
question_3_data = read_csv("./data_processed/cps_industry_and_demographic_data.csv")

vtable(question_1_data)
vtable(question_2_data)
vtable(question_3_data)
```

