---
title: "Analysis: COVID Effect on Employment"
author: "Group 2"
date: "`r Sys.Date()`"
knit: (function(input, ...) {rmarkdown::render(input, output_file="../paper/analysis.html")})
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Research Questions

* How has COVID affected the health of the retail industry, as measured by employment?
* How has retail fared relative to other industries?
* Retail needs to worry about who has money to spend - what has changed about who is      working and earning money?

Assumptions
A adult (over) civilian labor force. That the severe stay-at-home lock downs happened on roughly from 2022-03-01 until 2022-05-30.

## Libraries

``` {r libraries, echo=FALSE}
# Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
library(vtable)
library(fixest)
library(tidyverse)
library(lubridate)
library(marginaleffects)
library(grid)
```

## Read in Data

Bring in the processed data and store it in the schools variable. Filter on predominately bachelor degree schools. View the data.

```{r data read}
data = read_csv("./data_processed/cps_industry_and_demographic_data.csv")

```
# Seperate Data for Questions 1 and 2
```{r data split}
# This data can be used to answer question 2 and 1 at the YEARMONTH / INDNAME level.
industry_data = data %>% group_by(YEAR, MONTH, YEARMONTH, INDNAME, EMPCOUNTSTD, AFTERCOVID) %>% summarise()

# This dataset can answer question 1.
retail_data = industry_data %>% filter(INDNAME == "Retail Trade") 
```

# Question 1: 
# How has COVID affected the health of the retail industry, as measured by employment?
```{r}
ggplot(retail_data, aes(x = YEARMONTH, y = EMPCOUNTSTD)) + geom_point() + geom_line() + geom_smooth(method = 'lm') + geom_vline(xintercept = as.Date("2020-03-01")) + xlab("Date (Year and Month)") + ylab("Standardized Employment Count") + ggtitle("Standardized Employment Count by Year and Month for the Retail Industry")
```